\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-30 NAR_DRAFT.tex
check in

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
\begin{equation}
  \label{eqn:hypothesis}
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
\end{equation}
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

@@

Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e. $\mathrm{p}(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
\begin{equation}
  \label{eqn:2}
$$
  P(m,\nu) = p(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
$$
\end{equation}
Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.

@@

\begin{equation}
  \label{eqn:detailed_prior}
  $$
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{p(M)}{p(C = c)},
  \end{array}
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S}_{m,c})$ is the probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

@@

The number of bases $N$ is fixed at $3\cdot10^9$ for a whole genome sequence and $3\cdot10^7$ for a whole exome.
Returning to the set of high confidence mutations $\mathrm{D}$, we can set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, $f_{min}$ as the lowest, and count the number of mutations in $D$ falling between the two.
In practice it is usually best to select a conservative value for $f_{min}$, which is strongly dependent on sequencing depth, and we include this as a parameter in the R implementation of \batcave.
In all results presented in this paper we have set $f_{min}$ to 0.05 because we are working with very high sequencing depth.
A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in \citet{Williams2016}.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death, so that $\beta = 1$, which for most plausible values of $\beta$ means we over-estimate $\mu$ by a small amount.
Equation \ref{eqn:mut_rate} then allows us to estimate $p(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.

@@

\begin{table}
  \tableparts{%
  \caption{Performance characteristics as measured for all data sets.}
  \label{table:01}%
  }{%
  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
  \toprule
  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
  \\
  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
  \\
  \colrule
  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
  \\
  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
  \\
  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
  \\
  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
  \\
  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
  \\
  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
  \\
  \citet{Griffith2015} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
  \\
  \citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
  \\
  \botrule
  \end{tabular*}%
  }
  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
 \end{table}

>>>>

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector \bkmcomment{push equation onto separate line} $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
\begin{equation}
  \label{eqn:hypothesis}
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
\end{equation}
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

@@

Assuming that the identity of the alternate allele and its allele frequency are independent of each other, and $\nu$ is uniformly distributed (i.e. $\mathrm{p}(\nu) = 1$), we can rewrite equation \ref{eqn:1} as
\begin{equation}
  \label{eqn:2}
$$
  \mathrm{P}(m,\nu) = \mathrm{p}(m) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i).
$$
\end{equation}
Here we show how \batcave can be used to provide a tumor and site-specific estimate of the prior probability of mutation $\mathrm{p}(m)$.

@@

\begin{equation}
  \label{eqn:detailed_prior}
  $$
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)},
  \end{array}
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S}_{m,c})$ is the probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

@@

The number of bases $N$ is fixed at $3\cdot10^9$ for a whole genome sequence and $3\cdot10^7$ for a whole exome.
Returning to the set of high confidence mutations $\mathrm{D}$, we can set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, $f_{min}$ as the lowest, and count the number of mutations in $D$ falling between the two.
In practice it is usually best to select a conservative value for $f_{min}$, which is strongly dependent on sequencing depth, and we include this as a parameter in the R implementation of \batcave.
In all results presented in this paper we have set $f_{min}$ to 0.05 because we are working with very high sequencing depth.
A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in \citet{Williams2016}.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death, so that $\beta = 1$, which for most plausible values of $\beta$ means we over-estimate $\mu$ by a small amount.
Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.

@@

\begin{table}[b]
  \tableparts{%
  \caption{Performance characteristics as measured for all data sets.}
  \label{table:01}%
  }{%
  \begin{tabular*}{\textwidth}{@{}lllllllll@{}}
  \toprule
  Depth & Mutation profile & $\mu$ & AUROC & & AUPRC & & ICI & 
  \\
  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
  \\
  \colrule
  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
  \\
  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
  \\
  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
  \\
  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
  \\
  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
  \\
  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
  \\
  \citet{Griffith2015} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
  \\
  \citet{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
  \\
  \botrule
  \end{tabular*}%
  }
  {$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
 \end{table}
